---
title: "stringsAsFactors = HELLNO"
author: "Junbin ZHANG"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{stringsAsFactors = HELLNO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Factors are a very useful type of variable in R, but they can also drive you nuts. Especially the "stealth factor" that you think of as character.

Can we soften some of their sharp edges?

# Load `foofactors`

```{r}
library(foofactors)
```


# Retrieve information from factor

Factors contain a set of character values, and it uses a concept called "levels" for storing a pre-defined set values. When you are playing around factors, you may want to know what characters it contain as well as its levels. This package provides functions `fdetect()` for you to check if a factor can be considered as a character, and `freq_out()` to output frequency of each level as a table,

## `freq_out()`

Often we want a table of frequencies for the levels of a factor. The base `table()` function returns an object of class `table`, which can be inconvenient for downstream work. Processing with `as.data.frame()` can be helpful but it's a bit clunky.

```{r}
# complicted way to get frequencies of each levels
set.seed(1234)
x <- factor(sample(letters[1:5], size = 100, replace = TRUE))
table(x)
as.data.frame(table(x))
```

The `freq_out()` function returns a frequency table as a well-named `tbl_df`:

```{r}
# simply use freq_out() to get frequencies of each levels
freq_out(x)
```

## `fdetect()`

The `fdetect()` function checks if a factor can be considered as a character or not, based onn the fact that a character should have a length equals to its unique values, while a factor should not.

```{r}
# create a true factor and a character factor
true_factor <- factor(c("s", "t", "a", "t"))
character_factor <- factor(c("s", "t", "a"))
                      
# return TRUE if it is a true factor
fdetect(true_factor)

# return FALSE if it is a character factor
fdetect(character_factor)
```

# Modify factor and its levels

This package also provides functions for you to modify factors and, more importantly, control their levels. We provide `fbind()` for you to combine two factors, and `freorder()` and `fset()` to arrange levels.

## `fbind()`

Let's try `fbind()` with two factors.

```{r}
# create two factors to combine
a <- factor(c("character", "hits", "your", "eyeballs"))
b <- factor(c("but", "integer", "where it", "counts"))
```

Simply catenating two factors leads to a result that most don't expect.

```{r}
# wrong example to combine two factors
c(a, b)
```

The `fbind()` function glues two factors together and returns factor.

```{r}
# correct example to combine two factors with fbind()
fbind(a, b)
```

## `freorder()`

The `freorder()` function can reorder the levels of a factor in descending order. Let's have an example on `gapminder` to show the effects of it.

```{r fig.width = 7}
# load gapminder and tidyverse
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))

# original order
gapminder %>%
  # group by continent and country
  group_by(continent, country) %>% 
  # fill with continent
  ggplot(aes(x = continent, fill = continent)) +
  # make it a bar plot
  geom_bar() + 
  # change axis labels
  xlab("Continents") +
  ylab("Count") +
  # add title
  labs(title = "Number of countries in each continents") +
  # change theme
  theme_bw()
```

We can reverse the levels of continents using `freorder()`. As a funny example, let's reorder the color!

```{r fig.width = 7}
# reversed order
gapminder %>%
  # group by continent and country
  group_by(continent, country) %>% 
  # fill with reversed levels of continent
  ggplot(aes(x = continent, fill = freorder(continent))) +
  # make it a bar plot
  geom_bar() + 
  # change axis labels
  xlab("Continents") +
  ylab("Count") +
  # add title
  labs(title = "Number of countries in each continents") +
  # change theme
  theme_bw()
```


## `fset()`

The `fset()` function sets the levels of a factor by the order of apperance of its data. Let's try its effects on colors in the above example!

```{r fig.width = 7}
# reversed order
gapminder %>%
  # group by continent and country
  group_by(continent, country) %>% 
  # fill with reversed levels of continent
  ggplot(aes(x = continent, fill = fset(continent))) +
  # make it a bar plot
  geom_bar() + 
  # change axis labels
  xlab("Continents") +
  ylab("Count") +
  # add title
  labs(title = "Number of countries in each continents") +
  # change theme
  theme_bw()
```



# File I/O for data frame

This package provides read/write functions for you to handle your data frames. In addition, since this is a package for factors, these functions can help you retain factor levels in your data frame!

## `dfwrite()`/`dfread()`

These two functions save data frame as plain text delimited files (CSV files) while retaining factor levels.

```{r, message=FALSE}
# create a data frame as example
df <- data.frame(
    x = 1,
    y = 1:10,
    fac = sample(LETTERS[1:3], 10, replace = TRUE)
  )

# show the original levels of fac
levels(df$fac)

# reorder the levels of fac
df$fac <- freorder(df$fac)
levels(df$fac)

# write and read
dfwrite(df, "./examples/df.csv", "./examples/df_levels.txt")
read_df <- dfread("./examples/df.csv", "./examples/df_levels.txt")

# confirm levels are maintained
levels(read_df$fac)
```
