---
title: "stringsAsFactors = HELLNO"
author: "Junbin ZHANG"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{stringsAsFactors = HELLNO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Factors are a very useful type of variable in R, but they can also drive you nuts. Especially the "stealth factor" that you think of as character.

Can we soften some of their sharp edges?

# `fbind()`

Binding two factors via `fbind()`:

```{r}
library(foofactors)
a <- factor(c("character", "hits", "your", "eyeballs"))
b <- factor(c("but", "integer", "where it", "counts"))
```

Simply catenating two factors leads to a result that most don't expect.

```{r}
c(a, b)
```

The `fbind()` function glues two factors together and returns factor.

```{r}
fbind(a, b)
```

Often we want a table of frequencies for the levels of a factor. The base `table()` function returns an object of class `table`, which can be inconvenient for downstream work. Processing with `as.data.frame()` can be helpful but it's a bit clunky.

```{r}
set.seed(1234)
x <- factor(sample(letters[1:5], size = 100, replace = TRUE))
table(x)
as.data.frame(table(x))
```

# `freq_out()`

The `freq_out()` function returns a frequency table as a well-named `tbl_df`:

```{r}
freq_out(x)
```

# `fdetect()`

The `fdetect()` function checks if a factor can be considered as a character or not, based onn the fact that a character should have a length equals to its unique values, while a factor should not.

```{r}
# create a true factor and a character factor
true_factor <- factor(c("s", "t", "a", "t"))
character_factor <- factor(c("s", "t", "a"))
                      
# return TRUE if it is a true factor
fdetect(true_factor)

# return FALSE if it is a character factor
fdetect(character_factor)
```

# `freorder()`

The `freorder()` function can reorder the levels of a factor in descending order.

```{r}
# create a factor as an example
f <- factor(c("s", "t", "a"))

# the originally order of level should be "a" "s" "t"
levels(f)

# the reordered levels should be "t" "s" "a"
levels(freorder(f))
```

# `fset()`

The `fset()` function sets the levels of a factor by the order of apperance of its data.

```{r}
# create a factor as an example
f <- factor(c("s", "t", "a"))

# the originally order of level should be "a" "s" "t"
levels(f)

# the reset levels should be "s" "t" "a"
levels(fset(f))
```

# `dfwrite()`/`dfread()`

Functions `dfwrite()` and `dfread()` only work on data frame. `dfwrite()` saves data frame as plain text delimited files (CSV files) while retaining factor levels. Notive that:

- `dfwrite()` will generate a companion file, so both of these two functions require two generated files to work.
- Only columns of factor type in the resultant data frames can be reconstructed. For other columns, they may not be the same as the original ones because plain text delimited files drop many information.

In the following examples, the generated files are under ["examples"](./examples)

```{r, message=FALSE}
# create a data frame as example
df <- data.frame(
    x = 1,
    y = 1:10,
    fac = sample(LETTERS[1:3], 10, replace = TRUE)
  )

# show the original levels of fac
levels(df$fac)

# reorder the levels of fac
df$fac <- freorder(df$fac)
levels(df$fac)

# write and read
dfwrite(df, "./examples/df.csv", "./examples/df_levels.txt")
read_df <- dfread("./examples/df.csv", "./examples/df_levels.txt")
# confirm levels are maintained
levels(read_df$fac)
```

If no path is specified to store levels, a "levels.txt" file will be generated automatically under the same folder as the CSV file for data frame.

```{r, message=FALSE}
# write and read
dfwrite(df, "./examples/df.csv")
read_df <- dfread("./examples/df.csv", "./examples/levels.txt")
# confirm levels are maintained
levels(read_df$fac)
```
